Methods for Analyzing Models of Rod-Shaped Bacteria

### 2025-07-18 paper/main.tex
insert table here
mark sections with TODO
extend introduction with more references and bullet points

%###################################################################################################
\section{Introduction}
\label{section:introduction}

Bacteria exist in many physical shapes such as spheroidal, rod-shaped or spiral
\cite{Zapun2008,Young2006}.
The last decades have fundamentally challenged and changed our understanding of how bacteria retain
their shape.
Rod-shaped bacteria can grow by extending their circular part, or by inserting new material at the
tip of the rod.
Species such as E.coli and B.subtilis~\cite{Errington2020} fall under the first category while
S.pombe represents the latter.
New material is inserted in small bursts on the nanoscale in forms of patches, bands or
hoops~\cite{DePedro2003}.
Despite their differences in wall-size, B.subtilis and E.coli follow growth rules which are
comparable with respect to the extension of the rod~\cite{Chang2014}.

@@

%###################################################################################################
\subsection{Visualization}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

@@

% --------------------------------------------------------------------------------------------------
\subsubsection{Relationship to Semantic Segmentation}
\label{paragraph:comparison-semantic-segmentation}
\cite{Jadon2020}

@@

\begin{table}
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material along the circular part of the rod
            \cite{Robert2014,Takeuchi2005}.\\
        % (3) & Differentiation &
        %     \textit{B.subtilis} is known to differentiate into matrix-producing and
        %     surfactin-producing cells \cite{vanGestel2015,Lpez2010}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod, thus forming two new
            agents.
            Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (4) & Variable Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (5) & Adhesion &
            Bacteria adhere to each other at longer distances and attach when in close contact
            \cite{Verwey1947,Trejo2013}.\\
            % The interaction between rods can be polarized \cite{Duvernoy2018}.\\
        % (7) & Friction &
        %     Friction between cells \cite{Grant2014} can be asymmetrical \cite{Doumic2020}.\\
        &\textbf{(DC) Domain-Cell Interactions}\\
        \midrule
        (6) & External Forces &
            Bacteria which rub against surfaces tend to stick to them after some
            time~\cite{vanLoosdrecht1989}.
            The extracellular gel which surrounds the bacteria exerts a force onto the cells
            \cite{Grant2014}.\\
        % (9) & Extracellular Reactions &
        %     Bacteria can take up nutrients or possible secrete/take up signalling molecules
        %     \cite{Li2025}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{TODO Insert caption}
\end{table}

@@

%###################################################################################################
\section{Discussion}
\label{section:discussion}

@@

\section{Conclusion}
\label{section:conclusion}

>>>>

%###################################################################################################
\section{Introduction (TODO)}
\label{section:introduction}

Bacteria exist in many physical shapes such as spheroidal, rod-shaped or spiral
\cite{Zapun2008,Young2006}.
The last decades have fundamentally challenged and changed our understanding of how bacteria retain
their shape.
Rod-shaped bacteria can grow by extending their circular part, or by inserting new material at the
tip of the rod.
Species such as E.coli and B.subtilis~\cite{Errington2020} fall under the first category while
S.pombe represents the latter.
New material is inserted in small bursts on the nanoscale in forms of patches, bands or
hoops~\cite{DePedro2003}.
Despite their differences in wall-size, \textit{B.subtilis} and \textit{E.coli} follow growth rules
which are comparable with respect to the extension of the rod~\cite{Chang2014}.

\paragraph{Experimental studies of mechanical properties}
\textbf{Study of Individual Bacteria}
\begin{itemize}
    \item \cite{Robert2014} growth of individual rods, distribution of parameters
    \item \cite{Koutsoumanis2013} inheritance of growth rates, growth parameter distributions
    \item \cite{Ursell2014} MreB acts againts curvature during growth of cell
    \item \cite{Takeuchi2005} cells in microchambers leads to plastic deformations
    \item \cite{Amir2014} bending stiffness, plastic + elastic deformations
\end{itemize}
\textbf{Collective Effects \& Self-Organization}
\begin{itemize}
    \item \cite{Trejo2013} mechanical response of bacterial colony
    \item \cite{Duvernoy2018} bacterial colonies form multilayers, interactions are polar
    \item \cite{vanGestel2015} van Gogh bundles form from interplay of 2 cell-types
\end{itemize}

% --------------------------------------------------------------------------------------------------
\subsection{Numerical Models of Rod-Shaped Bacteria}
\textbf{Requirements}
\begin{itemize}
    \item capture bendiness; do not assume rigid bodies
    \item model in 2D and 3D possible
    \item individual-based model; model single cells
    \item collective phenomena; describe many interacting cells, not just a single one
\end{itemize}

\textbf{Computational Modeling Frameworks}
What has been done so far?
\begin{itemize}
    \item In general: not many frameworks which support some form of rod-shaped bacteria
    \item \cite{Kang2014} Biocellion has support for cylindrically-shaped agents; no
        bendiness/rigidity or polar interactions
    \item \cite{Matyjaszkiewicz2017} BSim2.0 represents cells as rigid capsular cells made from a
        cylindrical center part and two half-spheres, which are placed at the ends of the cylinder
        to round out the shape. In order to calculate interactions between cells, possible overlaps
        are determined and minimized, thus determining the position values of the next iteration
        step.
        BSim does not consider bending forces for individual cells or polar interactions.
    \item \cite{Gutirrez2017} The gro programming language was designed to simulate the growth of
        colonies and cell-cell communication. Its “physics computation has been optimized for rigid
        rod-shaped bodies, like E. coli bacteria” (Gutiérrez et al. 2017). They recognize two types
        of forces which are acting on the cellular agents: Local forces which are calculated between
        adjacent bacteria and a global force which pushes bacteria outwards of the colony. The
        latter of these is a phenomenological implementation of the observed colony expansion and
        the associated central pressure with it. This assumption may yield incorrect results for
        sparsely populated cases.
        The engine is limited to 2D and does not consider polar interactions or bending of the rods.
\end{itemize}

% --------------------------------------------------------------------------------------------------
\subsection{Purpose-built Models}
\begin{itemize}
    \item \cite{Abar2017} Review of various ABMs
    \item \cite{Winkle2017} Modeling mechanical interations in growing populations of rod-shaped
        bacteria
    \item \cite{Doumic2020} "A purely mechanical model with asymmetric features for early
        morphogenesis of rod-shaped bacteria micro-colony"
    \begin{itemize}
        \item Really good paper for referencing
        \item comparison of distributions; not individual agents
        \item no bending in model
        \item uses steric force (see also \cite{Trejo2013})
    \end{itemize}
    \item \cite{Grant2014} purposely-built model written in C++
    \begin{itemize}
        \item describes bacteria as collection of overlapping spheres
        \item spheres are coupled by non-linear springs (Euler-Bernoulli dynamic beam theory). This
            assumption is werid but does not alter their results
        \item Only model repulsive forces; no attraction, adhesion
    \end{itemize}
    \item \cite{Cho2007} only 2D model; no param estimation; based on work done in \cite{Jnsson2005}
    \item \cite{Storck2014} no bending; 41 parameters for various cases; only 8 parameters taken
        from literature values or quantified; generate growth rate from normal distribution at each
        integration step; stochastically equivalient for collection of bacteria but unfounded for
        individual bactera
\end{itemize}

%###################################################################################################
\section{Overview (TODO)}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Visualization (TODO)}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

@@

% --------------------------------------------------------------------------------------------------
\subsubsection{Relationship to Semantic Segmentation (TODO)}
\label{paragraph:comparison-semantic-segmentation}
\cite{Jadon2020}

@@

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material along the circular part of the rod
            \cite{Robert2014,Takeuchi2005}.\\
        % (3) & Differentiation &
        %     \textit{B.subtilis} is known to differentiate into matrix-producing and
        %     surfactin-producing cells \cite{vanGestel2015,Lpez2010}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod, thus forming two new
            agents.
            Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (4) & Variable Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (5) & Adhesion &
            Bacteria adhere to each other at longer distances and attach when in close contact
            \cite{Verwey1947,Trejo2013}.\\
            % The interaction between rods can be polarized \cite{Duvernoy2018}.\\
        % (7) & Friction &
        %     Friction between cells \cite{Grant2014} can be asymmetrical \cite{Doumic2020}.\\
        &\textbf{(DC) Domain-Cell Interactions}\\
        \midrule
        (6) & External Forces &
            Bacteria which rub against surfaces tend to stick to them after some
            time~\cite{vanLoosdrecht1989}.
            The extracellular gel which surrounds the bacteria exerts a force onto the cells
            \cite{Grant2014}.\\
        % (9) & Extracellular Reactions &
        %     Bacteria can take up nutrients or possible secrete/take up signalling molecules
        %     \cite{Li2025}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{TODO Insert caption}
\end{table}

@@

%###################################################################################################
\section{Discussion (TODO)}
\label{section:discussion}

@@

\section{Conclusion (TODO)}
\label{section:conclusion}
